Portable Multi- and Many-Core Performance for Finite Difference Codes; Application to the Free-Surface Component of NEMO.

### 2017-07-10 nemolite2d_perf.tex
nemolite2d-paper: update to copernicus 4.6 package and add line-numbering

%% 2-column papers and discussion papers
\documentclass[gmd, manuscript]{copernicus}

@@

Although we have investigated how the performance of the {PS}y{KA}l
version of NEMOLite2D compares with that of the original, we have not
addressed how efficient the original actually is. In order to do so we
use the Roofline Model~\citep{roofline} which provides a relatively
simple way of characterising the performance of a code in terms of
whether it is memory-bandwidth bound or compute bound. To do so we
follow the approach suggested in~\cite{para_pearls} and use the
STREAM~\citep{stream} and LINPACK~\citep{linpack} benchmarks in order to
get the upper bounds on the memory bandwidth and floating-point
operations per second (FLOPS) for the test hardware. Since we are
using an Intel Ivybridge CPU we used the Intel Math Kernel Library
implementation of LINPACK.

A key component of the Roofline model is the Operational or Arithmetic
Intensity of the code being executed:
\[
AI = \frac{\textrm{No. of floating-point operations}}{\textrm{Bytes fetched from memory}}
\]
We calculated this quantity manually by examining the
source code and counting the number of memory references and
arithmetic operations that it contained. In doing this counting we
assume that any references to adjacent array elements (\textit{e.g.}
$u(i,j)$ and $u(i+1,j)$) are fetched in a single cache line and thus
only count once.

@@

These performance estimates are plotted as CPU ceilings (horizontal
dashed lines) in Figure~\ref{FIG_roofline}. The performance of the
Momentum kernel is seen to fall between these two bounds which
demonstrates that its performance is not memory-bandwidth limited, as
might have been assumed by its low AI. Therefore, although the
performance of this kernel is clearly sub-optimal, this is due to the
balance of floating-point operations that it contains. Improving this
is outside the scope of the current work since here we are focused on
the {PS}y{KA}l separation of concerns and the introduction of parallelism
in the PSy layer.

Enabling SIMD vectorisation for this kernel does not significantly
improve its performance (Figure~\ref{FIG_roofline}) and in fact,
limiting it to SSE instructions (vector width of two double-precision
floating point numbers) rather than AVX (vector width of four) was
found to produce a slightly more performant version. The poor
efficiency of the SIMD version is highlighted by the fact that the
corresponding kernel performance no longer falls within the bounds of
our performance estimate. This is because we have incorporated the
effect of SSE into that estimate by simply assuming perfect
vectorisation which gives a performance increase of a factor of two.

>>>>

%% 2-column papers and discussion papers
\documentclass[gmdd, manuscript]{copernicus}

\usepackage{lineno}
\linenumbers

@@

Although we have investigated how the performance of the {PS}y{KA}l
version of NEMOLite2D compares with that of the original, we have not
addressed how efficient the original actually is. Without this
information we have no way of knowing whether further optimisations
might yield worthwhile performance improvements. Therefore, we
consider the performance of the original, serial NEMOLIte2D code on
the Intel Ivy Bridge CPU and use the Roofline Model~\citep{roofline}
which provides a relatively simple way of characterising the
performance of a code in terms of whether it is memory-bandwidth bound
or compute bound. To do so we follow the approach suggested
in~\cite{para_pearls} and construct a roofline model using the
STREAM~\citep{stream} and LINPACK~\citep{linpack} benchmarks in order
to obtain appropriate upper bounds on the memory bandwidth and
floating-point operations per second (FLOPS), respectively. Since we
are using an Intel Ivy Bridge CPU we used the Intel Math Kernel Library
implementation of LINPACK.

A key component of the Roofline model is the Operational or Arithmetic
Intensity (AI) of the code being executed:
\[
AI = \frac{\textrm{No. of floating-point operations}}{\textrm{Bytes fetched from memory}}
\]
We calculated this quantity manually by examining the
source code and counting the number of memory references and
arithmetic operations that it contained. In doing this counting we
assume that any references to adjacent array elements (\textit{e.g.}
$u(i,j)$ and $u(i+1,j)$) are fetched in a single cache line and thus
only count once.

@@

These performance estimates are plotted as CPU ceilings (horizontal
dashed lines) in Figure~\ref{FIG_roofline}. The performance of the
Momentum kernel is seen to fall between these two bounds which
demonstrates that its performance is not memory-bandwidth limited, as
might have been assumed by its low AI. Therefore, although the
performance of this kernel is well below the peak performance of the
CPU, this is due to the balance of floating-point operations that it
contains. Improving this is outside the scope of the current work
since here we are focused on the {PS}y{KA}l separation of concerns and
the introduction of parallelism in the PSy layer.

Enabling SIMD vectorisation for this kernel does not significantly
improve its performance (Figure~\ref{FIG_roofline}) and in fact,
limiting it to SSE instructions (vector width of two double-precision
floating point numbers) rather than AVX (vector width of four) was
found to produce a slightly more performant version. The poor
efficiency of the SIMD version is highlighted by the fact that the
corresponding kernel performance no longer falls within the bounds of
the performance estimates produced by Habakkuk. This is because we
have incorporated the effect of SSE into that estimate by simply
assuming perfect vectorisation which gives a performance increase of a
factor of two.  Further investigation of this issue revealed that
several of the NEMOLite2D kernels make frequent use of floating-point
division operations. Although SSE and AVX versions of the division
operation are available, on Ivy Bridge they do not provide any
performance benefit~\citep{fog_instructions}. A straightforward
optimisation then would be to alter the kernels in order to reduce the
number of division operations. However, again, kernel optimisation is
outside the scope of this paper since it breaks the PSyKAl separation
of concerns.
